---
title: "Assignment Local Poisson regression"
author: "Joan Oliveras Torra, Odysseas Kyparissis, Louis Tichelman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```
## 1. Bandwidth choice for the local Poisson regression

We modify the functions h.cv.sm.binomial and loglik.CV.

More precisely we only modify loglik.CV, in order to use leave-one-out CV using local Poisson regression. 

```{r}
library(sm)
# Bandwidth choice in the local Poisson regression 
# by leave-one-out cross-validation.
# Function "sm.poisson", from library "sm", is used.
#
# Pedro Delicado

# method can be equal to 'loglik.CV' (default) or 'prob.missclas.CV'
h.cv.sm.poisson <- function(x,y,rg.h=NULL,l.h=10,method=loglik.CV){
   cv.h <- numeric(l.h)
   if (is.null(rg.h)){
      hh <- c(h.select(x,y,method="cv"),
              h.select(x,y,method="aicc"))#,hcv(x,y))
      rg.h <- range(hh)*c(1/1.1, 1.5)
   }
   i <- 0
   gr.h <- exp( seq(log(rg.h[1]), log(rg.h[2]), l=l.h))
   for (h in gr.h){
      i <- i+1
      cv.h[i] <- method(x,y,h)
   }
   return(list(h = gr.h, 
               cv.h = cv.h, 
               h.cv = gr.h[which.min(cv.h)]))
}

# method loglik.CV: leave-one-out log-likelihood
loglik.CV <- function(x,y,h){
  n <- length(x)
  pred <- sapply(1:n,
      function(i,x,y,h){
         sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
      },   x,y,h)
  return(-sum(log(exp(-pred)*pred^(y)/factorial(y)))/n)
}
```

## 2. Local Poisson regression for Country Development Data

Attach the HDI data and create le.fm.r.

```{r}
countries<-read.csv2(file="HDI.2017.subset.csv",row.names = 1)
attach(countries)
countries$le.fm.r <- round(le.fm)
head(countries)
attach(countries)
```

We call the function h.cv.sm.poisson in order to obtain a bandwidth for the local Poisson regression based on the leave-one-out cross-validation.

We will set as h range from 1 to 10.

```{r}
library(sm)
# using h.CV.loglik
h.CV.loglik <- h.cv.sm.poisson(Life.expec,le.fm.r,rg.h=c(1,10),method=loglik.CV)
best.h<-h.CV.loglik$h[which.min(h.CV.loglik$cv.h)]

plot(h.CV.loglik$h,h.CV.loglik$cv.h, xlab = 'bandwidth', ylab = 'loo-CV estimation of (negative) expected log-likelihood', main = 'bandwidth choice')
lines(h.CV.loglik$h,h.CV.loglik$cv.h)
points(best.h, min(h.CV.loglik$cv.h), col = 2, pch = 16)
```

The best bandwidth parameter h according to LOOCV is:

```{r}
print(best.h)
```

Therefore, we fit a local Poisson regression modeling le.fm.r as a function of Life.expec. using the best value of bandwidth found.

```{r}
best.mod <- sm.poisson(Life.expec,le.fm.r,h=best.h,col=1)
```













