---
title: "Smoothing and Regression Splines"
author: "Joan Oliveras Torra, Odysseas Kyparissis, Louis Tichelman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

The file `bikes.Washington.Rdata` contains information on the bike-sharing rental service in Washington D.C., USA, corresponding to years `2011` and `2012`. This file contains only one data frame, `bikes`, with `731` rows (one for each day of years 2011 and 2012, that was a leap year) and `9` columns:

```{r}
load("bikes.Washington.Rdata")
summary(bikes)
```

From the summary and a brief inspection of the dataset the following information is obtained:

-   `instant`: Row index, ranging from 1 to 731.
-   `yr`: Year (0: 2011, 1: 2012).
-   `dayyr`: Day of the year (1 to 365 for 2011, 1 to 366 for 2012).
-   `weekday`: Day of the week (0 for Sunday, 1 for Monday, ..., 6 for Saturday).
-   `workingday`: If the day is neither a weekend nor a holiday, it is 1; otherwise, it is 0.
-   `temp`: Temperature in Celsius.
-   `hum`: Humidity in percentage (%).
-   `windspeed`: Wind speed in miles per hour.
-   `cnt`: Count of total rental bikes. In this assignment, we consider this variable as continuous.

# Question 1: Cubic regression spline

The model we want to estimate is: ($cnt \sim m(instant)$)

## Model fitting

```{r}
# Sort dataframe based on the "instant" column
df <- bikes[order(bikes$instant), ]
spline_fit <- smooth.spline(df$instant, df$cnt, cv = FALSE)
spline_fit
```

## Extraction of results

```{r}
lambda <- spline_fit$lambda
df <- spline_fit$df
num_knots <- spline_fit$fit$nk
fitted_values <- predict(spline_fit, bikes$instant)$y
```

## Display of results

### (a), (b), (c)

```{r}
# Display the results
cat("a) The chosen penalty parameter λ is:", lambda, "\n")
cat("b) The corresponding equivalent number of degrees of freedom df is:", df, "\n")
cat("c) The number of knots used is:", num_knots, "\n")
```

### (d) Graphic representation of the estimated regression function

The estimated regression function $\hat m(instant)$, together with the scatter plot of the real data is displayed in this section.


```{r}
plot(x = bikes$instant, y = bikes$cnt,
     main=paste("Number of knots =",num_knots,
                "; Equiv. no. params.=",round(df,2)), col = "purple", pch = 20)
abline(v=spline_fit$fit$min+spline_fit$fit$knot*spline_fit$fit$range, col=8, lty=2)
lines(spline_fit,col="orange",lwd=2)
# Add legend with adjusted position
legend("topleft", legend = c("Real Values", "Estimated Regression Function"), col = c("purple", "orange"), pch = c(20, NA), lwd = c(NA, 2), cex = 0.8)
```

# Question 2: Nonparametric logistic regression using splines with a IRWLS procedure

The following script (`IRWLS_logistic_regression.R`) includes the implementation of `logistic regression` by using a custom `Iteratively Re-Weighted Least Squares (IRWLS)` algorithm.

```{r}
# Source IRWLS_logistic_regression.R
source("IRWLS_logistic_regression.R")
```

The implementation is mainly performed from the function `logistic.IRWLS.splines` where its arguments are the explanatory variable $x$, the $0-1$ response variable $y$, the vector $x.new$ of new values of variable $x$ where we want to predict the probability of $y$ being $1$ given that $x$ is equal to $x.new$, the equivalent number of parameters (or model degrees of
freedom) $df$, and the logical `plts` indicating if plots are desired or not.

## Creation of variable cnt.5000

```{r}
bikes$cnt.5000 <- ifelse(bikes$cnt >= 5000, 1, 0)
```

The following `barplot` provides an insight on the number of days with more or less than $5000$ rentals.

```{r}
# Plot the counts of each category using a bar plot
barplot(table(bikes$cnt.5000), main = "Counts of cnt.5000 Categories", xlab = "cnt.5000", ylab = "Count", col = c("purple", "orange"), legend = c("0", "1"))
```

## (a) Range of temperatures for probability of having more than 5000 rentals in a day is larger than 0.5

```{r}
# Fit non-parametric binary regression
# Sort dataframe based on the "temp" column
df <- bikes[order(bikes$temp), ]
logistic_spline_fit <- logistic.IRWLS.splines(x = df$temp, y = df$cnt.5000, df = 6, plts=TRUE)

```

```{r}
# Calculate the estimated probabilities for the given temperatures
predicted_probabilities <- logistic_spline_fit$fitted.values
```



```{r}
# Find the temperatures where Pr(cnt >= 5000|temp) is larger than 0.5
temperatures <- df$temp[predicted_probabilities > 0.5]
```


```{r}
# Display the range of temperatures
cat("The range of temperatures where Pr(cnt >= 5000|temp) is larger than 0.5 is: [",min(temperatures),",",max(temperatures), "]\n")
```
## (b) k-fold log-likelihood cross validation

```{r}
# Set of possible values for df
df.v <- seq(3, 15)

# Initialize variables to store cross-validation results
cv_results <- numeric(length(df.v))

# Perform k-fold log-likelihood cross-validation
set.seed(789)  # Set seed for reproducibility

# Shuffle the indices
indices <- sample(1:nrow(bikes))

# Number of folds
k <- 5

# Size of each fold
fold_size <- ceiling(nrow(bikes) / k)

# Create list to hold fold indices
folds <- vector("list", k)

# Split indices into folds
for (i in 1:k){
  start <- ((i - 1) * fold_size) + 1
  end <- min(i * fold_size, nrow(bikes))
  folds[[i]] <- indices[start:end]
}

for (j in 1:length(df.v)){
  df = df.v[j]
  cv_log_likelihood = 0
  for (i in 1:k){
    x = bikes$temp[-folds[[i]]]
    x.new = bikes$temp[folds[[i]]]
    y = bikes$cnt.5000[-folds[[i]]]
    y.new = bikes$cnt.5000[folds[[i]]]
    model = logistic.IRWLS.splines(x=x, y=y, x.new = x.new, plts = F, df=df)
    pred = model$predicted.values
    valid_log_likelihood  = sum(y.new*log(pred) + (1-y.new)*log(1-pred))
    cv_log_likelihood = cv_log_likelihood + valid_log_likelihood
  }
  cv_results[j] = cv_log_likelihood
}

# Choose the df value that maximizes cross-validated log-likelihood
optimal_df <- df.v[which.max(cv_results)]
```


```{r}
# Display the cv_results data
cv_results
# Display the optimal df value
cat("Optimal df:", optimal_df, "\n")
```

```{r}
plot(df.v, cv_results, main = "Degrees of freedom (df) vs Average CV log-likelihood", xlab = "df", ylab = "Average CV log-likelihood",  col = "purple", pch = 20)
abline(v = optimal_df, col = "orange")  # Assuming optimal_df is a vertical line
```


```{r}
# Sort dataframe based on the "instant" column
df <- bikes[order(bikes$temp), ]
# Fit logistic regression model with optimal df
final_model <- logistic.IRWLS.splines(x = df$temp,
                                       y = df$cnt.5000,
                                       df = optimal_df,
                                       plts=TRUE)
```


```{r}
plot(x = df$temp, y = df$cnt.5000,
     main=paste("Degrees of freedom=",round(optimal_df,2)), col = "purple", pch = 20)
# lines(bikes$temp,final_model$fitted.values>0.5,col="orange",lwd=2)
lines(df$temp,final_model$predicted.values,col="orange",lwd=2)
# Add legend with adjusted position
legend("left", legend = c("Real Values", "Estimated Logistic Function"), col = c("purple", "orange"), pch = c(20, NA), lwd = c(NA, 2), cex = 0.5)
```

